---
title: "Data 608 - Module 3"
author: "Gabriel Santos"
date: "2023-03-04"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
runtime: shiny
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(shiny)
```


```{r}
data <- read.csv('https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv')
```

```{r}
summary(data)
```

```{r}
head(data)
```

Exploring the data:

Subset for 2010 only
calculate national average of crude mortality rate
National average for each disease

```{r}
data <-data[!(data$State==''),]
data <- data[complete.cases(data), ]
colnames(data) <- c('Disease', 'State', 'Year', 'Deaths', 'Population', 'Crude.Rate')
data$Disease <- as.character(data$Disease)
data$Crude.Rate <- as.numeric(data$Crude.Rate)
data$Disease <- substr(data$Disease, regexpr('>', data$Disease)+1, nchar(data$Disease))

data_2010 <- data[data$Year == 2010, ]

temp <- data_2010 %>% group_by(Disease)  %>% summarise(total_population = sum(Population)) %>% inner_join(data_2010, by="Disease")
temp$Weighted_CrudeRate <- (temp$Population/temp$total_population) * temp$Crude.Rate

data_2010 <- 
temp %>% group_by(Disease) %>% summarise(disease_national_avg = sum(Weighted_CrudeRate)) %>% inner_join(data_2010, by="Disease")
national_avg_2010 <- as.data.frame(unique(data_2010 %>% group_by(Disease) %>% select(Disease, disease_national_avg)))
```


Question 1: As a researcher, you frequently compare mortality rates from particular causes across different States. You need a visualization that will let you see (for 2010 only) the crude mortality rate, across all States, from one cause (for example, Neoplasms, which are effectively cancers). Create a visualization that allows you to rank States by crude mortality for each cause of death.


```{r echo=FALSE, fig.width=12, fig.height=12}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput(inputId = "disease", 
                  label = "Select disease:",
                  choices = unique(data_2010$Disease),
                  selected = "Neoplasms",
                  width = '500px')
    ),
    
    mainPanel(
      plotOutput(outputId = "bargraph")
    )
  )
)
server <- function(input, output, session) {
  
    ordered <- eventReactive(
    eventExpr = input$disease, 
    valueExpr = {data_2010 %>% filter(Disease==input$disease) %>% arrange(desc(Crude.Rate))},
    ignoreNULL = FALSE
  )
   
  output$bargraph <- renderPlot({ggplot(data = ordered(), aes(x=reorder(State,Crude.Rate), y=Crude.Rate)) + 
  geom_bar(stat="identity", width=0.7, color="#142336", fill="lightskyblue", 
           position=position_dodge()) +
    geom_text(aes(label=round(Crude.Rate, digits=2)), hjust=1.3, size=3.0, color="black") + 
    coord_flip() + 
    ggtitle("Crude Mortality  \n Each cause of death for 2010") +
    xlab("") + ylab("") + 
    theme_minimal()}, height = 1200, width = 700)
}
shinyApp(ui = ui, server = server)
```

For example: According to the visualization, we can see that for 2010 the state that presented the highest deaths due to Neoplasms was West Virginia (WV). Followed by Maine (ME).


Question 2: Often you are asked whether particular States are improving their mortality rates (per cause) faster than, or slower than, the national average. Create a visualization that lets your clients see this for themselves for one cause of death at the time. Keep in mind that the national average should be weighted by the national population.

```{r echo=FALSE, fig.width=12, fig.height=12}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput(inputId = "disease", 
                  label = "Select disease:",
                  choices = unique(data_2010$Disease),
                  selected = "Neoplasms",
                  width = '500px')
    ),
    
    mainPanel(
      plotOutput(outputId = "bargraph")
    )
  )
)
server <- function(input, output, session) {
  
  ordered <- eventReactive(
    eventExpr = input$disease, 
    valueExpr = {data_2010 %>% filter(Disease==input$disease) %>% arrange(Crude.Rate)},
    ignoreNULL = FALSE
  )
  
  national_avg <- eventReactive(
   eventExpr = input$disease,
   valueExpr = {unlist(national_avg_2010 %>% filter(Disease==input$disease) %>% select(disease_national_avg))}
  )
  
  output$bargraph <- renderPlot({ggplot(data = ordered(), aes(x=reorder(State,-Crude.Rate), y=Crude.Rate)) + 
  geom_bar(stat="identity", width=0.7, color="#142336", fill="lightskyblue", 
           position=position_dodge()) +
     
    geom_hline(yintercept=national_avg(), color="red", size=1) + 
    ggtitle("Crude Mortality for each cause of death for 2010 \n National Average") +
    xlab("") + ylab("") + 
    theme_minimal()}, height = 600, width = 900)
}
shinyApp(ui = ui, server = server)
```

For example: According to the visualization, we can see that for 2010, 23 states had mortality rates higher than the national average for deaths from Neoplasms.

